Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models

### 2021-12-15 variable_predictions.Rnw
Autocommit

set.seed(9991)
load("variable_predictions_objs.rda")
load("cubic_predictors_preds.rda")
load("cubic_predictors_preds_adjust.rda")
load("glme_random_intercept_preds.rda")
@

@@

Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s). In the former case, we treat the interactions as just another column in the variable space (of the model matrix). In the later case, the non-focal variables in the model matrix are averaged as before. However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations. For example, consider the model~\ref{eq:simple_inter_higher} below: 
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
\end{align}
%
For the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is 
%
\begin{align*}
\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\bar{x}_2\bar{x}_3
\end{align*}
%
where $x_{1i}$ are the carefully chosen levels of the focal predictor. On the other hand, for the second case, if we consider $x_2$ as the focal predictor, then the interaction ($x_2x_3$) is between focal and non-focal predictor. In this case, the linear predictor is given by 
%
\begin{align*}
\eta^\star(x_{2ij}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 \bar{x}_1 + \beta_2x_{2ij} + \beta_3x_{3j} + \beta_{23}x_{2ij}x_{3j}
\end{align*}
%
In general, our formulation, even for more complicated interactions, follow these two basic principles -- interaction between non-focal predictors and interaction between focal and non-focal predictors.

@@

In Figure~\ref{fig:pred_cubic_plots} we compare the non-isolated (everything) and the isolated predictor effects. In the first case (Figure~\ref{fig:pred_cubic_plots}a), the focal predictor, \code{age}, has a higher order interaction in the form of cubic polynomial while in the second case (Figure~\ref{fig:pred_cubic_plots}b), the focal predictor, \code{wealthindex}, is simple but the non-focal predictor is a cubic polynomial. In Figure~\ref{fig:pred_cubic_plots}a, the average prediction and the ``truth'' are very similar but in this case, cubic polynomial, the model center is not a point in the predictor space. On the hand, in Figure~\ref{fig:pred_cubic_plots}b, the average prediction is higher than the observed marginal, ``truth'' but the model center is a point in the predictor space. 
%
\begin{figure}[H]
\begin{center}
<<pred_cubic_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(pred_cubic_plots)
@
\end{center}
\caption{A comparison of predictor effects and their corresponding CIs. In Figure a) the focal predictor has a higher cubic polynomial interaction while in Figure b) the focal predictor has no interaction but the non-focal predictor is a cubic polynomial. The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively. The vertical grey line is the mean of the focal predictor (model center). The dotted black (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects together with their corresponding CIs.}
\label{fig:pred_cubic_plots}
\end{figure}
%

Even in simple interaction model in Figure~\ref{fig:pred_age_cont_plots}b and Figure~\ref{fig:pred_cubic_plots}b, we start to observe bias in the predicted and observed means. To illustrate the bias correction, we regenerate the predictor effects in Figure~\ref{fig:pred_cubic_plots}b but incorporate bias adjustment.
%
\begin{figure}[H]
\begin{center}
<<pred_cubic_plots_adjust, echo=FALSE, fig=TRUE, message=FALSE>>=
print(pred_cubic_plots_adjust)
@
\end{center}
\caption{A comparison of bias corrected and uncorrected predictor effects. The red solid and dotted lines are not bias corrected (none), the black ones are corrected for bias, the horizontal grey and yellow lines are the observed and predicted average household size, respectively, while the vertical dotted line is the average wealth index. In the absence of bias, the two approaches give similar estimates (Figure a). On the other hand, in Figure b, without correcting for the bias, the predictions are slightly higher than the corrected ones which matches the observed relatively well.}
\label{fig:pred_cubic_plots_adjust}
\end{figure}
%

@@

Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household and we are interested in the status improved water services (improved or unimproved). In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation such that 
%
\begin{align}\label{sim:lme_one_pred}
\mathrm{logit(status} = 1) &= \eta \nonumber\\
\mathrm{\eta} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
i &= 1,\cdots, 500
\end{align}
%
We compare predictor effects generated without any bias correction to those corrected for the bias, and then compare the average predicted probability to the observed. The results are shown in Figure~\ref{fig:glme_plots}. The uncorrected approach overestimated the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability. 
%
\begin{figure}[H]
\begin{center}
<<glme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(glme_plots)
@
\end{center}
\caption{A compare of bias corrected and uncorrected predictor effects. The red lines are the bias uncorrected estimates while the black ones are the bias corrected. The horizontal red and black lines are average predicted probability of improved water. The horizontal grey (overlaid by the black) dotted line is the observed average probability of improved water. The vertical dotted grey line is the average age while the grey points are the binned observations.}
\label{fig:glme_plots}
\end{figure}
%

>>>>

set.seed(9991)
load("variable_predictions_objs.rda")
load("cubic_predictors_preds.rda")
load("cubic_predictors_preds_adjust.rda")
load("glm_one_predictor_preds.rda")
load("glme_random_intercept_preds.rda")
load("binom_multiple_outcomes_preds_plots.rda")
@

@@

Higher order terms such as interactions, splines, polynomials, etc., can be between the non-focal predictors or focal and non-focal predictor(s). In the former case, we treat the interactions as just another column in the variable space (of the model matrix). In the later case, the non-focal variables in the model matrix are averaged as before. However, for the interacting focal predictors, a combination of each unique levels (or quantiles) are first generated and then the interaction terms are generated by multiplying these combinations. For example, consider the model~\ref{eq:simple_inter_higher} below: 
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
\end{align}
%
For the first case, if we consider $x_1$ as the focal predictor, the interaction ($x_2x_3$) is between the non-focal predictors, $x_2$ and $x_3$, the linear predictor is 
%
\begin{align*}
\eta^\star(x_{1i}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 x_{1i} + \beta_2\bar{x}_2 + \beta_3\bar{x}_3 + \beta_{23}\over{x_2x_3}
\end{align*}
%
where $x_{1i}$ are the carefully chosen levels of the focal predictor. On the other hand, for the second case, if we consider $x_2$ as the focal predictor, then the interaction ($x_2x_3$) is between focal and non-focal predictor. In this case, the linear predictor is given by 
%
\begin{align*}
\eta^\star(x_{2ij}, \nset{{\bar{x}^\star}}) = \beta_0\mathbf{1} + \beta_1 \bar{x}_1 + \beta_2x_{2ij} + \beta_3x_{3j} + \beta_{23}x_{2ij}x_{3j}
\end{align*}
%
In general, our formulation, even for more complicated interactions, follow these two basic principles -- interaction between non-focal predictors and interaction between focal and non-focal predictors.

@@

In Figure~\ref{fig:pred_cubic_plots} we compare the non-isolated (everything) and the isolated predictor effects. In the first case (Figure~\ref{fig:pred_cubic_plots}a), the focal predictor, \code{age}, has a higher order interaction in the form of cubic polynomial while in the second case (Figure~\ref{fig:pred_cubic_plots}b), the focal predictor, \code{wealthindex}, is simple but the non-focal predictor is a cubic polynomial. In both cases, the average prediction and the ``truth'' are very similar but in Figure~\ref{fig:pred_cubic_plots}a, cubic polynomial focal predictor, the model center is not a point in the predictor space. On the hand, in Figure~\ref{fig:pred_cubic_plots}b, the model center is a point in the predictor space. 
%
\begin{figure}[H]
\begin{center}
<<pred_cubic_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(pred_cubic_plots)
@
\end{center}
\caption{A comparison of predictor effects and their corresponding CIs. In Figure a) the focal predictor has a higher cubic polynomial interaction while in Figure b) the focal predictor has no interaction but the non-focal predictor is a cubic polynomial. The dotted yellow and grey horizontal lines are the expected and observed/true means, respectively. The vertical grey line is the mean of the focal predictor (model center). The dotted black (non-focal uncertainty excluded) and blue lines (all uncertainty included) are the predictor effects together with their corresponding CIs.}
\label{fig:pred_cubic_plots}
\end{figure}
%

\subsection{Generalized linear model}

Consider model described in \ref{sim:lm_no_interaction} with only \code{Age} as the predictor such that 
%
\begin{align}\label{sim:glm_one_pred}
\mathrm{logit(status} = 1) &= \eta \nonumber\\
\mathrm{\eta} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i
\end{align}
%
where \code{status} is a binary outcome with 0 for unimproved and 1 for improved water quality in the slums. By having only a single predictor, we may be able to investigate the contribution of nonlinear link function on the predicted probability of improved water quality.

Figure~\ref{fig:glm_plots} shows the predictor effect of \code{age} on predicted probability of improved water quality. For such a simple generalized linear model, there is no much bias induced by the nonlinear averaging induced by the nonlinear link function. In particular, the predicted average probability of improved water quality, horizontal black line, closely match the observed proportion of households with improved water quality, horizontal solid black line.
%
\begin{figure}[H]
\begin{center}
<<glm_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(glm_plots)
@
\end{center}
\caption{Predictor effect of age on improved water quality. The horizontal black and grey lines are the average predicted probability and observed proportion of improved water quality, respectively. The vertical dotted grey line is the mean age while the red points are the observed bines of proportion of improved water quality.}
\label{fig:glm_plots}
\end{figure}
%

@@

Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household and we are interested in the status improved water services (improved or unimproved). In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation such that 
%
\begin{align}\label{sim:lme_one_pred}
\mathrm{logit(status} = 1) &= \eta \nonumber\\
\mathrm{\eta} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
i &= 1,\cdots, 500
\end{align}
%
We compare predictor effects generated without any bias correction to those corrected for the bias, and then compare the average predicted probability to the observed. The results are shown in Figure~\ref{fig:glme_plots}. The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected estimates which closely match the observations and the observed average probability. 
%
\begin{figure}[H]
\begin{center}
<<glme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(glme_plots)
@
\end{center}
\caption{A comparison of bias corrected and uncorrected predictor effects. The red lines are the bias uncorrected estimates while the black ones are the bias corrected. The horizontal red and black lines are average predicted probability of improved water. The horizontal grey (overlaid by the black) dotted line is the observed average probability of improved water. The vertical dotted grey line is the average age while the grey points are the binned observations.}
\label{fig:glme_plots}
\end{figure}
%

\subsubsection{Multivariate binary outcomes}

We now consider a bit more complicated model involving two outcome variables (services) -- status (improved or unimproved) of water and garbage services. In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict probability of improved water and garbage services based on age and wealth index. Each household has its own correlated effect on the intercepts (random-intercept). In addition, we add latent variable to account for unobserved correlations.
%
\begin{align}\label{sim:lme_multi_outcome}
\mathrm{services}_{{\mathrm{h}[i]\mathrm{s}}} &= \mathrm{Binomial}(P_{{\mathrm{h}[i]\mathrm{s}}}) \nonumber\\
\mathrm{logit}(P_{{\mathrm{h}[i]\mathrm{s}}}) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Latent}_{[i]\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
\epsilon_{[i]\mathrm{\text{\tiny{water}}}} &\sim \mathrm{Normal}(0, 2) \nonumber\\
\epsilon_{[i]\mathrm{\text{\tiny{garbage}}}} &\sim \mathrm{Normal}(0, 3) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 2 \nonumber\\
\beta_{\mathrm{W}} &= -3 \nonumber\\
\beta_{\mathrm{L\text{\tiny{water}}}} &= 1 \nonumber\\
\beta_{\mathrm{L\text{\tiny{garbage}}}} &= -5 \nonumber\\
i &= 1,\cdots, 100
\end{align}
%
Where:
\begin{itemize}
\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline log odds of a household having improved service
\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
\end{itemize}

%
\begin{figure}[H]
\begin{center}
<<binom_outcomes_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(binom_outcomes_plots)
@
\end{center}
\caption{TODO}
\label{fig:binom_outcomes_plots}
\end{figure}
%
